Management induced changes of soil organic carbon on global croplands

### 2020-10-06 SOCbudgetPaper/discussion.Rmd
Mayor update of everything

- The natural land representation and therefor also SOC initialization is lacking proper parametrization of both decay behaviour and natural input (e.g. nitrogen and lignin content of litterfall). As all results are valued against these potential natural SOC the absolute values and also the changes over time have to be used with caution. Especially in boreal zone and in very dry regions the natural land representation might be far off, due to the focus for the parametrization for cropped areas. 
The good agreement of the stock change factors $F^{SCF}$ with the Tier 1 default values might serve as a hint that the natural land representation is not totally off. Boreal and very dry zone might be uncropped, leaving the results on SOC changes on cropland untouch from the bad borreal land representation.

- This study as the IPCC guidelines suggested has limited her focus to the first 30 cm of the soil profile, leaving changes in the subsoil unnoticed. Nevertheless studies (see Don on tillage) have shown, the subsoil to be a game changer in evaluating total SOC losses or gains for no-tillage systems. It has been argued that for intensivly tilled soils, subsoil SOC is increasing due to the import of carbon rich topsoil to deeper soil layers. Following these argumentation SOC stocks in croplands might even be underestimated.

- As pointed out by Keel (2017), Smith (2019) the results might be highly sensitive to carbon input calculations more precisly to below and above ground residue carbon estimates derived from harvested quantities. It has been questioned, that below ground residues might increase with a fixed root:shot ratio (maybe specifically in high end farming systems (?)). Following this argumentation SOC results shown in this study might especially in high-yielding farming system (europe etc.) overestimate actual SOC stocks. (In general the sensitivity to and the uncertainty of certain parameters might be very high.)

<!--Tasks: 
Find better natural land representation and initalization. Using LPJmL values might be complex since the split up into SOC subpools is needed. Getting better parametrization for the CENTURY based approach for decay behaviour and better values for nitrogen and lignin of natural litter fall.  -->

\newpage

@@

>>>>

## Including agricultural management data changes the sign of the trend

This study provides an analysis on historic SOC stock changes on cropland. We determine the SOC trends on cropland compared to a potential natural vegeted world under the same historic climatic conditions ($SOC_{natveg}$) leading to a cummulative SOC emission since dawn of mankind of around 37 GtC in 2010. Whereas recent modelling estimates on global SOC emissions indicate an on going increase of SOC emissions (Sanderman et al, Pugh et al.), our study indicates that the global SOC gap might be closing slowly.

According to Sanderman et al. cummulative SOC emissions since beginning of human cropping activities have been at around 37 GtC for the first 30 cm of the soil with half of it attributed to grazing. It was also pointed out, that this results might be conservative estimates and low compared to experimental results, leaving our estimate of 37 GtC in 2010 for cropland emissions only in alignment with Sanderman et al. estimates considering the hugh uncertainties in modelling SOC on the global scale. 
Furtherore the results of Sanderman et al. calculated historical trends based on agricultural land expansion without considering SOC variations due to different management systems at all. In Pugh et al. management effects like tillage and residue recycling have been considered, but neither changes over time nor alignment to oberserved historical data like yields levels or no tillage areas were taken into account. Under this assumptions this study only found marginal effect of crop productivity and other management effects on cummulative SOC emissions.

Nevertheless our results show, when purely focusing on the past decades, the moderate global cropland expansion of around 11% between 1974 and 2010, is outweighted by improved agricultural yields ad practises. Moreover our sensitivity analysis indicades that 1. yield increases and with that the increase in residue biomass might play the most important role, followed by 2. enhanced residue recycling rates, 3. improved manure recycling (e.g. due to improved animal waste management system) and 4. the adoption of no tillage practises.

Modelling management effects on the global scale comes however with huge parametic and systematic uncertainties. One aspect pointed out by Keel (2017) as well as by Smith (2019) might be that carbon input calculations are highly sensitive to the choice of allometric functions determining below and above ground residue estimates from harvested quantities. Keel et al. questioned, that below ground residues might increase with a fixed root:shot ratio rather than being independend from productivity gains. Following this argumentation SOC results shown in this study might especially in high-yielding farming system overestimate actual SOC stocks. However according to our study above ground residue biomass recycling seem to contribute even more to overall SOC stocks due to higher input rates.

## Modeled management effect inline with default IPCC assumptions

To validated the effect of our modelled SOC stocks and stock changes under management, we compare our results to default IPCC stock changes factors (cite), which are based on measurement data for croplands (see \@ref(tab:SCFtable)).

```{r caption_SCFtable, echo=FALSE}
caption_SCFtable <- "This table shows IPCC 2006 as well as 2019 default (medium input) and low input regimes stock change factors $F_{SCF}$ without other subsystem consideration compared results of this study (SOC budget) for 1990 and 2010."
```

```{r SCFtable, out.width = "16cm", fig.cap = caption_SCFtable, echo=FALSE, fig.pos="H"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/TableSCF_comparison.png")
```

Our climate zone specific, aggregated $F_{SCF}$ correspond very well to the tier 1 estimation of IPCC, 2006 of $F_{SOC}$. For the tropical regions the assumptions changed notablly from the guidelines in 2006 to the update in 2019, leaving our results too low in comparison with IPCC, 2019. Considering yield gaps in mainly developing regions in the tropics the default assumption of medium input systems, might be an overestimation of actual SOC state. The additional effect of considering low inpout regimes in tropical regions can however not explain the full mismatch to IPCC 2019 values but account for at least 5-7% of it.

## SOC stocks inline with literature

The worlds SOC stock and its changes are highly uncertain (cite), which is visible by the wide range of global SOC stock estimates (see \@ref(tab:SOCtable)).

```{r caption_SOCtable, echo=FALSE}
caption_SOCtable <- "Modelled as well as data based estimation for global SOC stock (for 2010) in GtC for the first 30 cm of soil."
```

```{r SOCtable, out.width = "8cm", fig.cap = caption_SOCtable, echo=FALSE, fig.pos="H"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/TableSOC_comparison.png")
```

The global esimates of SOC stock by this study are on the lower edge compared to other modelled results or more data driven estimates. Looking on regional results (fig. SX in supplement), our estimates turn out to be in good agreement for most regions with the largest deviations for boreal areas. Considering that the model was parametrized for croplands, these mismatches are not superising, since the temperature effects on decomposition is most likly fundamentally different for permaforst soils than assumend in this study. To avoid that this bias influences our results, our study focusses exclusively on cropland soils, excluding most of the boreal zone. Morover, when focussing on SOC changes, pristine natural vegetated areas without human land management under the same historic climatic conditions cancel out in the calculation of SOC emissions.

However the natural land representation that is dominating the total SOC stocks of the world and with it also the SOC initialization is lacking proper parametrization of nitrogen and lignin content of litterfall. This leaves carbon inputs and decay behaviour for natural land rather uncertained. As all results are valued against these potential natural SOC the absolute values and also the emissions over time have to be used with caution. Especially in less forested areas the natural land representation might be off, due to parameterization assumptions of the natural litterfall. Nevertheless total SOC stocks are in a reasonable range and the stock change factors $F^{SCF}$ are in good agreement with the Tier 1 default values of 2006. In addition we preformed a sensitivity analysis for lignin and nitrogen parameterization of natural litterfall (see appendix). It shows that the general trend of decreasing SOC emissions is not altered even by rather unlikely parameter choices, that would overestimate natural SOC stocks strongly and therefor lead to high emission in the case of land conversion.

## Important short commings --> move to appendix

@@

-This study as the IPCC guidelines suggested has limited her focus to the first 30 cm of the soil profile, leaving changes in the subsoil unnoticed. Nevertheless studies (see Don on tillage) have shown, the subsoil to be a game changer in evaluating total SOC losses or gains for no-tillage systems. It has been argued that for intensivly tilled soils, subsoil SOC is increasing due to the import of carbon rich topsoil to deeper soil layers. Following these argumentation SOC stocks in croplands might even be underestimated.

### 2020-10-06 SOCbudgetPaper/introduction.Rmd
Mayor update of everything

Possible titles:

* How management changed soil organic carbon on global croplands

* Agricultural soil have lost 37 GtC topsoil carbon

* Linking agricultural management data to soil modeleling – a global approach

Soil Organic Carbon (SOC), the amount of organic carbon stored through the earth's soil, is the largest terrestrial carbon pool, exceeding the carbon in the atmospheric and biospheric pools multiple times (Bathjes). As such, even small changes in drivers of SOC may thus lead to substantial shifts in earth carbon cycle and influence the amount of CO2 in the atmosphere  (ref. permafrost melting). The specific amount of carbon stored in the soil is however uncertain, with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (Bathjes, 1996). The quality of SOC maps has markedly improved in recent decades, along with better understanding of the factors driving the global magnitude, distribution, and dynamics of SOC pools. Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles in this regard.  
Human intervention, including land cover change and land management, has however added a further driver to SOC change, which alters terrestrial carbon pools in much shorter time scales and is likely one of the most dominant driver of SOC changes on managed land today. Recent studies identify the anthropogenic SOC debt for the first meter of the soil profile at around 116 GtC (Sanderman et al.), compared to previous estimates between 60-130 GtC <!--look up --> (Lal, 2006). Other studies have focused more closely on spatially disaggregation of SOC changes via advanced digital soil mapping techniques (S-World; Stoorvogel 2, 2017) or better representation of biogeochemical processes within SOC dynamics  (Hararuk, 20XX). While providing a rough estimate of the order of magnitude of change, these studies lack a detailed consideration of land management, especial agricultural activities.  
Field-scale models (ref. Daycent, RothC, Ecosse, C-Tool) are able to capture these land mangement impacts by using detailed information on crop yield levels, fertilizer inputs and various other on-farm activities. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global extent remains a complex challenge.  
Our study combines an spatially explicit estimate of agricultural management data on the global level with a 3-pool SOC model parametrized for global croplands. This allows us to estimate SOC stock change factors, as well as organic carbon flow dynamics within the agricultural system. We thereby consider change in SOC caused by historical land cover change as well as of different agricultural mangement practices, including residue recycling, manure amendments, irrigation, and tillage. We thereby provide the first global, spatially explicit SOC loss maps that consider detailed agricultural management.  
This paper will first introduce in the methods section the basic concept of SOC dynamics as applied in this study. We continue with a detailed description of the global gridded management data used here, including crop production levels, residue recycling rates, manure amendments, and the adoption of irrigation and tillage practices. Lastly, we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines.  In our results section we focus on the SOC dynamics of global croplands by (1) analysing the spatially explicit distribution and depletion of SOC from 1975 to 2010, (2) comparing the climate zone specfic stock changes with default stock change factors from the IPCC Tier 1 approach and (3) quantify global agricultural carbon flows and stocks to compare the importance of various management aspects. Finally, we will discuss our findings -- including their implications for SOC model development -- and conclude with an outlook on the ability of SOC management to mitigate climate change and contribute to negative greenhouse gas emissions.

>>>>

Soil Organic Carbon (SOC), the amount of organic carbon stored through the earth’s soil, is the largest terrestrial carbon pool, exceeding the carbon in the atmospheric and vegetation pools multiple times (Batjes). As such, even small changes in drivers of SOC may thus lead to substantial shifts in earth carbon cycle and influence the amount of CO2 in the atmosphere (ref. permafrost melting). The specific amount of carbon stored in the soil is, however, uncertain, with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (Batjes, 1996; more). The quality of SOC maps has markedly improved in recent decades, along with better understanding of the factors driving the global magnitude, distribution, and dynamics of SOC pools. Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles to determine SOC variations over space and time.
Human intervention, including land cover change and land management, has however added a further driver to SOC change, which alters terrestrial carbon pools in much shorter time scales and is likely one of the most dominant driver of SOC changes on managed land today (cite). Recent studies identify the anthropogenic SOC debt for the first meter of the soil profile at around 116 GtC (Sanderman et al.), compared to previous estimates between 60-130 GtC  (Lal, 2006). Other studies have focused more closely on spatially disaggregation of SOC changes via advanced digital soil mapping techniques (S-World; Stoorvogel 2, 2017) or better representation of biogeochemical processes within SOC dynamics (Hararuk, 20XX). While providing a rough estimate of the order of magnitude of change, these studies lack a detailed consideration of land management, especially of agricultural activities.
Field-scale models (cite Daycent, RothC, Ecosse, C-Tool) are able to capture these land management impacts by using detailed information on crop yield levels, fertilizer inputs and various other on-farm measures. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global extent remains a complex challenge. Our study combines an spatially explicit estimate of agricultural management data on the global level with a 3-pool SOC model parametrized for global croplands. This allows us to estimate SOC stock change factors, as well as organic carbon flow dynamics within the agricultural system. We thereby consider change in SOC caused by historical land cover change as well as of different agricultural management practices, including residue recycling, manure amendments, irrigation, and tillage. We provide the first global, spatially explicit SOC loss maps that considers detailed agricultural management.
This paper will first introduce in the methods section the basic concept of SOC dynamics as applied in this study. We continue with a detailed description of the global gridded management data used here, including crop production levels, residue recycling rates, manure amendments, and the adoption of irrigation and tillage practices. Lastly, we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines. In our results section we focus on the SOC dynamics of global croplands by (1) analysing the spatially explicit distribution and depletion of SOC from 1975 to 2010, (2) comparing impacts of different management effects on SOC emissions and (3) quantify global agricultural carbon flows and stocks to compare the importance of various management aspects. Finally, we discuss and compare our findings – including their implications for SOC model development – and conclude with an outlook on the ability of SOC management to mitigate climate change and contribute to negative greenhouse gas emissions.

### 2020-10-06 SOCbudgetPaper/methods.Rmd
Mayor update of everything

Note that the absolution changes $\Delta SOC$ can be also interpreted as cummulative SOC emissions or as SOC depth of mankind; whereas relative changes $F^{SCF}$ can be denoted as stock change factors as defined within the IPCC guidelines of 2006 [@ipcc_2006_2006].

@@

To account for the distribution of tillage of the three tillage types specfied by the IPCC - full tillage, reduced tillage and no tillage -, we assume that all natural land and pastures are not tilled, whereas annual crops are under full and perennials under reduced tillage. This rule is based on the approach of [@porwollik_generating_2018], but neglects the (less common) adoption of no-tillage farming on agricultural land.

>>>>

Note that the absolution changes $\Delta SOC$ can be also interpreted as the SOC gap due to human cropping activits; whereas relative changes $F^{SCF}$ can be denoted as stock change factors as defined within the IPCC guidelines of 2006 [@ipcc_2006_2006]. Note that the SOC gap is to the negated cummulative SOC emission also refered to as SOC debt (cite Sanderman) of mankind.

@@

To account for the distribution of tillage to the three different tillage types specfied by the IPCC - full tillage, reduced tillage and no tillage -, we assume that all natural land and pastures are not tilled, whereas as default annual crops are under full and perennials under reduced tillage. Furthermore we assume no tillage in cropland cells specified as no tillage cell based on the historic global gridded tillage data set from Prowollik  [@porwollik_generating_2018]. Porwollik data set is extended by to the period of 1974-2010 by combining country-level data on conservation agriculture (ca) area values from AQUASTAT (FAO, 2016) and LUH2 crop areas together with the methodolgy of Porwollik to identify potential no tillage cells.

### 2020-10-06 SOCbudgetPaper/results.Rmd
Mayor update of everything

```{r table, echo = FALSE}
table <- readRDS("../ResultNotebooks/Output/SFC_climatezone.rds")
knitr::kable(table$naturalstate_table, "latex", digit=2, caption="SCF compared to potential natural state")
```

```{r table2, echo = FALSE}
table <- readRDS("../ResultNotebooks/Output/SFC_comparison.rds")
knitr::kable(table$comparison_table, "latex", digit=2, caption="SCF compared to IPCC Tier 1")
```

```{r out.width = "16cm"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/4panelfigure.png")
```

>>>>

We present simulation results of our SOC budget focusing on cropland areas for the year 2010 as well as global trends of SOC stock changes for the period from 1975 to 2010.

## SOC distribution and depletion

```{r caption_SOCmaps, echo=FALSE}
caption_SOCmaps <- "(a): Distribution of total global SOC stocks on cropland shows high carbon stocks in high yielding areas. (b): The SOC debt is decreasing over time, meaning net SOC gains on global croplands over the last decades. (c)+(d): Absolute (c) and relative (d) SOC stocks compared to a potential natural state showing different hot spots of SOC dynamics. Whereas the absolute losses might be in temperate dry regions, relative losses are more prominent in tropical moist areas."
```

```{r SOCmaps, out.width = "18cm", fig.cap = caption_SOCmaps, echo=FALSE, fig.pos="H"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/4panelfigure.png")
```

In fig. 1(a) we provide the first world map of SOC on croplands considering real world management data on the global scale. Our spatially explicit results moreover show hot spots of SOC losses as well as gains in two different ways: 
1. Absolute SOC changes (see fig. 1(c)) indicate areas with high importance for the global SOC emissions. The might be driven by huge relative losses or a high natural stock, from which even small deviations could lead to substantial losses.
2. To attribute SOC losses to insufficient agricultural management relative SOC changes ($F^{SCF}$, see fig. 1(d)) are a helpful tool. They indicate areas with huge difference in carbon inflows or SOC decay compared to natural vegetation, that might be overcome due to improved agricultural pratices.

## Agricultural management effects on SOC emissions and cycling

Global cummulative SOC emissions are decreasing (see fig. 1(c)). Fig. 2 reveals the relative impact of management effects by freezing tillage areas as well as carbon inflows from residues or manure at the level of 1975. Our counterfactual scenarios show that the increasing residue carbon input had the biggest overall effect on SOC stocks. Without changes in management regimes especially in residue inflows to the soil, global cummulative SOC emissions would still grow. The strong effect of carbon residue amounts are also visible in the carbon flow diagram for agricultural production for the year 2010 (see fig. 3)

```{r caption_SOCscen, echo=FALSE}
caption_SOCscen <- "Global SOC gap in GtC for various stylized management counterfactual scenarios compared to the modeled historical baseline."
```

```{r SOCscen, out.width = "18cm", fig.cap = caption_SOCscen, echo=FALSE, fig.pos="H"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/scenario.png")
```

```{r caption_FlowFig, echo=FALSE}
caption_FlowFig <- "Global carbon flows (small numbers) and stocks (bold numbers) within the agricultural system for the year 2010 (in MtC): Most important carbon sources on cropland are crop residues. Note the two numbers on carbon inputs to soil denote carbon applied to the field and carbon entering the soil (difference is decomposted before official counting as soil)."
```

```{r FlowFig, out.width = "16cm", fig.cap = caption_FlowFig, echo=FALSE, fig.pos="H"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/OuFlowFig.png")
```

\newpage
